Union and intersection contracts: theory vs practice

### 2021-03-23 paper.tex
Add citations and subsection on contracts

\documentclass{article}
\usepackage{mathtools,amssymb,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{array}
\usepackage[toc,page]{appendix}
\usepackage{graphicx}
\usepackage{varioref}
\usepackage{verbatim,comment}
\usepackage{longtable}
\usepackage{tikz}

@@

This paper is centered around the design of the Nickel language. Nickel is a
configuration programming language, meaning that a Nickel program eventually
evaluates to a value that can be directly serialized to JSON, YAML or alike.
Nickel aims at being lightweight, easy to write and easy to embed in a larger
system, to serve either directly as a native language to specify configuration,
or as a meta-language to generate configuration. The focus is first and foremost
practical: Nickel originated as a rethink of the language of the Nix package
manager, used to describe the biggest set of software packages ever built [ ],
and targets build systems and cloud deployment tools (Kubernetes, Terraform or
NixOps) as well.  While there's no hesitation in incorporating recent research
when it serves a well-defined purpose, the design decisions are guided by the
industrial use cases.

Our aim is that writing simple Nickel should not be more complex than writing
JSON or YAML. Unlike JSON, though, Nickel anticipates large configurations by
being both programmable and typed. In the configuration setting, there is a
singular dilemma with respect to typing: if your program is only run once on
fixed inputs, as are such programmable configurations, any type error will be
reported at run-time anyway. Why bother with the complexity of a static type
system? On the other hand, more and more software systems offload complexity to
configurations, such as cloud infrastructure, build system, or operating
systems, leading the corresponding configurations to become increasingly
complex, to the point where static types are beneficial. For reusable code —
that is, library functions — static types brings all the usual, widely
acknowledged benefits.

This dilemma is naturally solved by gradual typing [], an approach capable of
reuniting static typing and dynamic typing. Unlike gradually-typed language like
TypeScript though, we do not seek to build a complex type system that tries hard
to type most of the idioms naturally arising in dynamically typed code.  Rather,
we choose to provide a reasonably expressive type system with good inference
properties, that should make writing typed generic function that operate on data
relatively easy, but may require resorting to untyped code for more exotic expressions.
However, this is not a surrender, as in fact Nickel provides an alternative and
complementary mechanism for more complex data validation, contracts[references].

@@

In programming, unions are fundamental and ubiquitous building blocks of data,
and Nickel is no exception. While products represent conjunctions, which are several pieces of
data stored all together in the same value, unions represent disjunctions, which are
several exclusive alternatives stored in one value. When nested together, unions
and intersections are able to describe a large class of data shapes: lists,
trees, structured errors, and so on [ref, examples?].

\begin{verbatim}
EXAMPLES of union in configuration specification
\end{verbatim}

In statically typed languages, and more specifically in functional programming
languages, unions are usually embodied by algebraic data types [references],
also called \textbf{tagged unions}. Tagged means that there is an additional
field, or constructor, that allows to discriminate between each alternative.
When an union is constructed, the programmer specifies which alternative is
chosen by using the corresponding tag. When a union is consumed, usually
via pattern matching, the code branches depending on the tag, and each branch is
ensured to handle data of the right shape. Well-known programming languages with
first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
Swift and surely many more.

\begin{verbatim}
EXAMPLES
\end{verbatim}

@@

In the statically typed setting, the actual representation of a tagged union is
an implementation detail: from the point of view of the programmer, a tagged
union is a fist-class data type associated with constructors and destructors all
with well-defined semantics.  However, in a gradually typed language like
Nickel, the question of the representation of tagged unions as untyped values
arises.

@@

A better way to think of it is maybe as dependent $\Sigma$-type which first
component is a tag and second component is the data type corresponding to this
tag:
\[
    \Sigma_{tag : <A, B, C>} \text{Alternative}(\text{tag})
\]

Alas, choosing this representation as a default is not fully satisfying. As any
representation, it does not map precisely to a JSON type. In consequence, a lot
of illegal combinations are structurally valid as untyped JSON values. More
importantly, this does not necessarily match with the configuration schemas one
can find in the wild, that favor simple, untagged union [example of JSON
schemas/protobuf/whatever with unions].

In dynamically typed languages, which are interpreted for the vast majority
[ref/examples?], it is customary to directly store different kind of data in one
variable without further ceremony. The runtime system needs to attach type tags
to values anyway, which can usually be observed using functions like the
JavaScript's \verb+typeof+. In practice, using so-called \textbf{untagged
unions} has already been an idiom in dynamically typed language for a long time
[ref/examples], and being able to type and handle these unions is an important
aspect of the migration to a gradual type system. Indeed, most of gradually
typed languages either existed originally as dynamically typed language or aim
at backward compatibility with a dynamically typed base language. This is why
they tend to favor supporting untagged union types, the typed counterpart of
this practice [example of gradually typed language with untagged union].

In the case of Nickel, there is no pre-existing untyped language that we have to
support, so we are freer of our choice. But are we? As a configuration language,
Nickel does not evolve in the abstract. It aims at the best interoperability
possible with JSON. It is designed to generate data for other softwares to
consume, and tries as much as possible to be expressive enough to encode data
schemas at the language level. If these schemas use untagged union, then this is
a strong incentive to support them at the type level, or at least, at the
contract level.

@@

Encoding as: use singleton contract for tags, and then union.
\begin{lstlisting}
type sum = A of Num | B of Str
\end{lstlisting}

@@

Flow typing[ ], as implemented for example by TypeScript, is capable of solving
this issue by correctly propagating the type information in each branch. One can
imagine doing the same in Nickel. Thus equipped with untagged unions and a
typechecker smart enough, classic tagged unions can be simulated, but with the
benefit that the language is not imposing any representation. Untagged unions
empower the programmer to represent and handle a wider variety of unions
appearing in the wild. Thus, adding union types (and their natural and
unavoidable counterpart, intersections types [ref]) to Nickel sounds appealing.

@@

\section{Contracts}
\color{red}Goals of the section

A primer on contracts, which in the end are the problematic bit when confronted
with unions.\vspace{0.5cm}\color{black}

>>>>

\documentclass{article}
\usepackage{mathtools,amssymb,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{array}
\usepackage[toc,page]{appendix}
\usepackage{graphicx}
\usepackage{varioref}
\usepackage{verbatim,comment}
\usepackage{longtable}
\usepackage{tikz}
\usepackage{url}

\usepackage{biblatex}
\addbibresource{nickel.bib}

@@

This paper is centered around the design of the Nickel
language\cite{NickelRepo}. Nickel is a configuration programming language,
meaning that a Nickel program eventually evaluates to a value that can be
directly serialized to JSON, YAML or alike.  Nickel aims at being lightweight,
easy to write and easy to embed in a larger system, to serve either directly as
a native language to specify configuration, or as a meta-language to generate
configuration. The focus is first and foremost practical: Nickel originated as a
rethink of the language of the Nix package manager, used to describe one of the
biggest sets of software packages ever built\cite{repology}.
Nickel targets build systems and cloud deployment tools (Kubernetes, Terraform or
NixOps) as well.  While there's no hesitation in incorporating recent research
when it serves a well-defined purpose, the design decisions are guided by the
industrial use cases.

Our aim is that writing simple Nickel should not be more complex than writing
JSON or YAML. Unlike JSON, though, Nickel anticipates large configurations by
being both programmable and typed. In the configuration setting, there is a
singular dilemma with respect to typing: if your program is only run once on
fixed inputs and terminate, as are such programmable configurations, any type
error will be reported at run-time anyway. Why bother with the complexity of a
static type system? On the other hand, more and more software systems offload
complexity to configurations, such as cloud infrastructure, build system, or
operating systems, leading the corresponding configurations to become
increasingly complex, to the point where static types are beneficial. For
reusable code — that is, library functions — static types brings all the usual,
widely acknowledged benefits.

This dilemma is naturally solved by gradual typing\cite{Siek06gradualtyping}, an
approach capable of reuniting static typing and dynamic typing. Unlike
gradually-typed language like TypeScript, we do not seek to build a complex type
system that tries hard to type most of the idioms naturally arising in
dynamically typed code.  Rather, we choose to provide a reasonably expressive
type system with good inference properties, that should make writing typed
generic function that operate on data relatively easy, but may require resorting
to untyped code for more exotic expressions.  However, this is not a surrender,
as Nickel provides a complementary mechanism for more complex data validation,
contracts.

\subsection{Contracts}
\color{red}Goals of the section

A primer on contracts, which in the end are the problematic bit when confronted
with unions.\vspace{0.5cm}

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Fellleisen introduces a principled approach to run-time predicate checking that
nicely supports higher-order functions. It turns out these contracts are closely
related to the type casts introduced by gradual typing. In
\cite{WellTypedBlamed}, the authors precisely propose a system mixing gradual
typing and contracts \textit{à la Findler \& Felleisen}. Nickel adopts a similar
type system, with both statically typed terms, dynamically typed terms, and
first-class contracts. In this setting, each construct of the type system -
arrow types, record types, polymorphic types - must have a contract counterpart,
that is able to check at run-time the membership of a value to the corresponding
type.

\color{black}

@@

Unions are fundamental and ubiquitous building blocks of program data. Products
represent conjunctions, that is several pieces of data whith potentially
different shapes that are packed together in one value. Unions represent
disjunctions, which are several exclusive alternatives stored in one value. When
nested together, unions and intersections are able to represent aribrary
tree-like structures, allowing to model a large class of data.

Unions are also found in configuration, to represent nullable value in JSON,
plain enumerations, or more general union types when several representations are
supported.

In statically typed languages, and more specifically in functional programming
languages, unions are usually embodied by algebraic data types [references],
also called \textbf{tagged unions}. Tagged means that there is an additional
field, or constructor, that allows to discriminate between each alternative.
When an union is constructed, the programmer specifies which alternative is
chosen by using the corresponding tag. When a union is consumed, usually via
pattern matching, code branches are tagged as well, and each branch is ensured
to handle data of the right shape. Well-known programming languages with
first-class tagged unions include Haskell, OCaml, Rust, Scala, PureScript, Elm,
Swift and surely many more.

\begin{lstlisting}[caption={Example of a binary tree as a tagged union in OCaml}, captionpos=b]
type binary_tree =
  | Node of 'a binary_tree * 'a binary_tree
  | Leaf of int

let sum =
  let rec aux total = function
  | Leaf val -> val
  | Node(left,right) =>
    let left = aux total left in
    aux total right
\end{lstlisting}

@@

In the statically typed setting, the actual representation of a tagged union is
an implementation detail: from the point of view of the programmer, a tagged
union is a fist-class data type associated with constructors and destructors all
with well-defined semantics. In contrast, in a gradually typed language like
Nickel, the question of the representation of tagged unions as untyped values
arises.

@@

\lstinline+<A, B, C>+ is the type of a C-like enumeration that can take value
$A$, $B$ or $C$, while \lstinline+Dyn+ is the unknown dynamic type.

A better way to describe it would be as a dependent pair which first component
is a tag and second component is the data type corresponding to this tag, but
the language does not feature dependent types:
\[
    \Sigma_{tag : <A, B, C>} \text{Alternative}(\text{tag})
\]

Alas, choosing the first representation as a default is not fully satisfying.
First, it does not map precisely to a JSON type: in consequence, a lot of
illegal combinations are structurally valid as untyped JSON values. But this is
to be expected of any representation. More importantly, this does not
necessarily match with the configuration schemas one can find in the wild, that
favor simple, untagged union.

In dynamically typed languages, it is customary to directly store different kind
of data in one variable without further ceremony. The runtime system needs to
attach type tags to values anyway, which can usually be observed using functions
like JavaScript's \verb+typeof+. In practice, using so-called \textbf{untagged
unions} has already been an idiom in dynamically typed language for a long time
[ref/examples], and being able to type and handle these unions is an important
aspect of the migration to a gradual type system. Indeed, a lot of gradually
typed languages either existed originally as dynamically typed language or aim
at backward compatibility with a dynamically typed base language. This is why
they tend to favor supporting untagged union types, the typed counterpart of
this practice (as e.g in TypeScript \cite{TypeScriptUnions}).

In the case of Nickel, there is no pre-existing untyped language that we have to
support, so we are freer of our choice. But as a configuration language, Nickel
does not evolve in the abstract. It aims at the best interoperability possible
with JSON. It is designed to generate data for other softwares to consume, and
tries as much as possible to be expressive enough to encode data schemas at the
language level. If these schemas use untagged union, then this is a strong
incentive to support them at the type level, or at least, at the contract level.

@@

\begin{lstlisting}
type sum = A of Num | B of Str
\end{lstlisting}

@@

Flow typing[ ], as implemented for example by TypeScript, is capable of solving
this issue by correctly propagating the type information in each branch. One can
imagine doing the same in Nickel. Thus equipped with untagged unions and a
typechecker smart enough, classic tagged unions can be simulated, but with the
benefit that the language is not imposing any representation. Untagged unions
empower the programmer to represent and handle a wider variety of union idioms
appearing in the wild. Thus, adding union types (and their natural and
unavoidable counterpart, intersections types [ref]) to Nickel sounds appealing.

@@

\printbibliography
